\documentclass[11pt]{letter}
\usepackage[margin=2.5cm]{geometry}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\signature{ZuKang Xu\\Houmo AI, Nanjing, China\\
\href{mailto:zukang.xu@houmo.ai}{zukang.xu@houmo.ai}}

\date{\today}

\begin{document}

\begin{letter}{Editor-in-Chief\\
\textit{Software: Practice and Experience}\\
Wiley}

\opening{Dear Editor,}

I am pleased to submit the enclosed manuscript, ``\textit{Paper Agent}: A
Local-First Human-in-the-Loop Agentic Workspace for Controllable Academic
Writing,'' for consideration for publication in \textit{Software: Practice
and Experience}.

This paper presents \textit{Paper Agent}, a local-first software platform
that integrates AI assistance into the full academic writing workflow while
keeping the researcher in control. The system introduces three key
contributions: (1) a permission-aware AI interaction model (Chat, Agent,
Tools) that separates read-only discussion, reviewable edit proposals, and
controlled tool execution; (2) typed workflow pipelines with human
checkpoints as first-class gates; and (3) citation verification against
multiple scholarly databases (CrossRef, Semantic Scholar, OpenAlex). The
implementation comprises approximately 23,000 lines of source code,
validated by 227 automated tests with 100\% pass rate.

This work aligns with SPE's focus on practical software engineering
contributions. It addresses a real-world problem---the fragmentation of
AI-assisted academic writing across disconnected tools---with an implemented,
tested, and openly available system. The paper follows SPE's convention of
reporting on a complete system with attention to architecture, testing,
performance, and deployment considerations.

This manuscript has not been published and is not under consideration for
publication elsewhere. I have no competing interests to disclose. During
manuscript preparation, I used LLM tools (GPT-4o and Claude) for language
polishing and grammar correction only; I have reviewed all AI-assisted
content and take full responsibility for the final manuscript.

I look forward to your consideration.

\closing{Sincerely,}

\end{letter}
\end{document}
